Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex
Minor wordsmithing.

In this section, we present a
splitting algorithm for the source language \lang, specifying
precisely the algorithm and its correctness conditions.  
We begin by defining a masking relation that serves to extract stage~\bbone\ and
stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting must match. We then 
move on to the specification of the splitting algorithm itself.

@@

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value (3), and an $e_2$ corresponding to the stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.

To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.

@@

Splitting a stage \bbone\ term $e$ in \lang\
% $\colone e A$ 
yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$ in \lang,
we require that $e_1$ reduce to the tuple $(y,z)$,
where $y$ is identical to $\masko{\xi;v}$, and 
$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$. Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (called the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

The combined term of a {\tt case} executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
The precomputation result $z_i$ for this branch is injected and bundled with that of the predicate ($z_1$).   
$z_i$ is cased in the resumer to determine which branch's resumer should be executed.
{\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.

@@

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, is defined on the local structure of $e$.
In the terminal cases of
constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers that produce the original term. For
example, the integer constant \texttt{3} splits into the
precomputation \texttt{()} and resumer \texttt{\_=>3}.

For non-terminal terms~$e$ in \lang\ (except $\prev$
and $\pause$)
% recursively splitting
% subterms and generating precomputations $p_1, \ldots, p_n$ and
% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.
A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
\ur{This last point seems important.  It would be important to come
  back to it.}

>>>>

In this section, we present a
splitting algorithm for the source language \lang, specifying
precisely the algorithm and its correctness conditions.  
We begin by defining a masking relation that serves to extract stage~\bbone\ and
stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
move on to the specification of the splitting algorithm itself.

@@

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value \texttt{3}, and an $e_2$ corresponding to the stage~\bbtwo\ computation: \\
$\letin \yhat {\mathtt{7+5}} \yhat$.

To precisely specify the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.

@@

Splitting a stage \bbone\ term $e$ in \lang\
% $\colone e A$ 
yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$ in \lang,
we require that $e_1$ reduce to the tuple $(y,z)$,
where $y$ is identical to $\masko{\xi;v}$, and 
$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$. Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (a value we call the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
The precomputation result $z_i$ for this branch is injected and bundled with that of the predicate ($z_1$).   
$z_i$ is cased in the resumer to determine which branch's resumer should be executed.
{\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.

@@

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, is defined on the local structure of $e$.
In the terminal cases of
constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers consisting of the original term.
For example, the integer constant \texttt{3} splits into the
precomputation \texttt{()} and resumer \texttt{\_=>3}.

For all non-terminal terms~$e$ in \lang\ (except $\prev$
and $\pause$)
% recursively splitting
% subterms and generating precomputations $p_1, \ldots, p_n$ and
% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.

A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
\ur{This last point seems important.  It would be important to come
  back to it.}
